Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-22 report/MSc_Report.Rnw
.

\chapter*{Abstract}
% TODO: add abstract text.

\chapter*{Acknowledgments}
% TODO: add acknowledgments.

>>>>

\chapter*{Abstract}
% Checklist (Abstract): [x] integrate supplied summary [ ] confirm PDF rebuild

This thesis studies multivariate density estimation for tabular data within a transport-based evaluation framework.
All models standardize features and learn a monotone lower-triangular map $S \colon \mathbf{u} \to \mathbf{z}$ to a Gaussian reference distribution.
The estimators accumulate exact log-Jacobians in standardized space with an affine correction for reporting on the data scale.
We compare separable triangular transport maps (TTM-Sep), additive-predictor Transformation Random Forests (TRTF), and copula baselines under the shared preprocessing pipeline.
Additive TRTF induces the same separable triangular likelihood as TTM-Sep via the probability integral transform.
The resulting estimator yields exact likelihoods, transparent conditionals, and linear per-sample cost but restricts context-dependent shape.

Benchmarks include synthetic Half-Moon and a four-dimensional autoregressive generator, plus the UCI POWER, GAS, HEPMASS, and MINIBOONE datasets.
For Half-Moon with $n = 250$, joint negative log-likelihoods (nats) were $1.71$ for TRTF, $1.93$ for TTM-Sep, and $1.54$ for the copula baseline.
For the four-dimensional generator, TRTF achieved $4.53$, TTM-Sep $5.66$, and the copula baseline $5.45$.
Permutation averages confirmed ordering sensitivity for triangular maps.
On MINIBOONE with $K = 43$ and $N = 2{,}500$, TRTF reached $-30.01$ test log-likelihood, whereas flow baselines remained near $-11.7$.

Findings indicate that TRTF narrows the gap to oracle likelihoods on low-dimensional synthetic data.
The method trails modern flows on high-dimensional real data because separability limits context-dependent shape.
Copula baselines remain competitive in low dimensions.
This thesis records guidance and limitations, including ordering effects and fragility in sparse regimes.
